---
title: "Session 4 Exercises"
author: "Dhruvi Sheth"
date: "12 Jan 2026"
output:
  pdf_document: default
  html_notebook: default
  word_document: default
---

```{r}
set.seed(42)
```

### Exercise 1: Normal Distribution

```{r , fig.width=5, fig.height=5}
weights <- rnorm(200, mean = 70, sd = 10)

hist(weights,
     probability = TRUE,
     col = "lightblue",
     main = "Body Weight Distribution",
     xlab = "Weight (kg)")

x <- seq(min(weights), max(weights), length = 100)
lines(x, dnorm(x, mean = 70, sd = 10), col = "red", lwd = 2)

```

```{r}
# Calculate mean and sd
cat("mean =", mean(weights), "\n")
cat("sd   =", sd(weights), "\n")
```


### Exercise 2: Binomial Distribution

```{r}
# Probability of infection = 0.2
# 50 individuals per experiment
infection <- rbinom(100, size = 50, prob = 0.2)

hist(infection,
     probability = TRUE,
     col = "lightgreen",
     main = "Number of Infected Individuals",
     xlab = "Count")
```

### Exercise 3: Poisson Distribution

```{r, fig.width=5, fig.height=5}
# Average mutations per gene = 1.5
mut <- rpois(200, lambda = 1.5)

hist(mut,
     probability = TRUE,
     col = "cyan",
     main = "Mutations per Gene",
     xlab = "Mutations")
```

### Exercise 4: Interpretation

1. Which distribution is symmetric?
Ans: Normal Distribution

2. Which one is skewed?
Ans: Poisson  is right-skewed.

3. Which biological scenario fits each distribution?
Ans: 

-Normal distribution:
Continuous traits like body weight, height, enzyme activity

-Binomial distribution:
Number of infected individuals, survival vs death, success/failure outcomes

-Poisson distribution:
Mutations per gene, RNA-seq read counts, rare events per unit time or space

