---
title: 'Session 2: Exercises'
author: "Dhruvi Sheth (22294)"
output:
  pdf_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

############################################################ 

## EXERCISE 1: Individuals and Variables

```{r}
data(iris)
```

#### Questions:

1. What is the individual?

2. How many variables are measured?

```{r}
str(iris)
```

#### Answers:

1.  Individual: Each individual is one iris flower (one row of the dataset).

2.  Number of variables: There are five variables measured.

############################################################ 

## EXERCISE 2: Classify Data Types

#### Question:

For each variable in iris, state whether it is:

- Continuous numerical

- Categorical

```{r}
str(iris)
```

-   Sepal.Length – Continuous numerical (num)
-   Sepal.Width – Continuous numerical (num)
-   Petal.Length – Continuous numerical (num)
-   Petal.Width – Continuous numerical (num)
-   Species – Categorical (Factor)

############################################################ 

## EXERCISE 3: Visualize Data Types

#### Questions:

1. Plot a histogram of Sepal.Width

```{r}
hist(iris$Sepal.Width,
     col = "lightgreen",
     main = "Sepal Width Distribution",
     xlab = "Sepal Width")
```

2. Plot a barplot of Species

```{r}
barplot(table(iris$Species),
col = c("pink", "cyan", "lightyellow"),
main = "Species Count",
ylab = "Frequency")

```

############################################################ 

## EXERCISE 4: Binary Data Simulation

#### Questions: 

1. Simulate infection status for 100 individuals
```{r}
set.seed(10)
infection <- sample(c("Infected", "Not Infected"),
                    size = 100,
                    replace = TRUE,
                    prob = c(0.3, 0.7))
```

2. Visualize the result

```{r}
barplot(table(infection),
col = c("orange", "blue"),
main = "Infection Status",
ylab = "Count")

```

############################################################ 

## EXERCISE 5: Observational vs Experimental

Scenario: You observe diet and diabetes status in a population.

#### Questions:

1. Is this observational or experimental?

2. What confounders might exist?

#### Answers:

1. This is an **observational study** because diet is not assigned.

2. Possible confounders that might exist:

-   Age
-   Genetics
-   Socioeconomic status
-   Body weight
-   Sleep
-   Amount of stress
-   Lifestyle
-   Medication and healthcare access

############################################################ 

## EXERCISE 6: Experimental Design

#### Question:
Design an experiment to test:
"Does fertilizer increase plant height?"

In comments, describe:

- Control group
- Treatment group
- Variable measured
- How you would randomize
- How would you have replicates

#### Answer:

- **Control group:** Plants that receive no fertilizer\
- **Treatment group:** Plants that receive fertilizer\
- **Variable measured:** Plant height\
- **Randomization:** Randomly assign plants to control or treatment groups to avoid bias 
- **Replication:** Use many plants per group (e.g., 20–30 plants) to reduce random errors.
